Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-03 appendix.tex
Cleaning out addressed comments.
Main text pass and half of appendix.

\subsection*{Site split formulation}
We begin by introducing ``site splits.''
One application of site splits is to formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
This is a standard step in the description of the Hadamard transform (Section 8.6 of \citet{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.

@@

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nSiteRows\times\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define the random variables
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
\]
\end{definition}
The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$.

@@

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
%EM Do discontinuities occur? In my mind we are taking the maximum of continuous functions, which is continuous.
%EM We don't think they occur.
%das: done
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
We show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.

All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms.
The corresponding possibilities for the conditional likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which shows the likelihood remains unchanged if $x_1$ and $x_2$ are swapped.
The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
Exchanging $x_1$ and $x_2$ also does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.

@@

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
%EM Perhaps we can say something about how this is not a strong assumption.
%EM see below "By the assumption of"
%das: moved below
For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
\label{lemma:bounds}
\end{lemma}

@@

%EM Also here.
%das: reworded
Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
We now prove the main result.

@@

%EM I agree that this condition shows that the lower bound is increasing on the entire interval. But how do we get to a conclusion about $\ell(.)$ itself?
%das: The derivative of \ell, i.e., \ell', is a decreasing function on [0,1]. The value of \ell' at 1 is positive, thus, \ell' > 0 on all of [0,1]. Since \ell' > 0 on all of [0,1], \ell is increasing on all of [0,1] (I think via MVT?).

>>>>

\subsection*{Site split formulation}
We begin by introducing ``site splits.''
We use site splits to formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
This is a standard step in the description of the Hadamard transform (Section 8.6 of \citet{Semple2003-em}), although our approach is complicated slightly by the inclusion of ancestral states.

@@

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nSiteRows\times\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=0,\\
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define the random variables
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
\]
\end{definition}
The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$ and so $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is $\overline{\ancestralStateColumn}$.

@@

\subsection*{Example}
We follow with an example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
%EM perhaps a section reference for this big book?
Using the notation in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.

All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, t) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
We show that for site splits $\{1\}$ and $\{3\}$, exchanging $x_1$ and $x_2$ also swaps the values of the conditional likelihood terms (Table~\ref{tab:farris_likelihoods}).
Indeed, the corresponding possibilities for the conditional likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, t) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which shows the likelihood remains unchanged if $x_1$ and $x_2$ are swapped.
The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$
%EM does this one need a different explanation now?
as well as $\{1,2\}$ and $\{2,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
Exchanging $x_1$ and $x_2$ also does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.

@@

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
\label{lemma:bounds}
\end{lemma}

@@

Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
We now prove the main result.

@@
